---
title: "final_data"
author: "Chris Hunter"
date: "2024-08-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(stringr)
library(janitor)
library(lubridate)


us_i <- clean_names(read_delim("data/us_confirmed.csv"))
us_d <- clean_names(read_delim("data/us_deaths.csv"))

#vax <- read_delim("data/us_state_vaccinations.csv")
lockdown <-clean_names(read_delim("data/countryLockdowndates.csv"))

# function to get new infection totals per month of a given year
get_new_cases_by_month <- function(year) {
  # month scale is absolute from Jan 2020. This gives each month a unique id so
  # we can aggregate year over year data
  add_months <- 12 * (as.numeric(year) - 2020)

  # grab this year's data for the selected states
  this_year <- us_i %>%
    filter(province_state %in% states) %>%
    filter(str_detect(date, year))

  # we'll add each month to this empty data frame in the loop below
  yearly_data <- data.frame()

  # calculate aggregate new cases per month for each state, and add to the
  # yearly data
  for(state in states) {
    this_state_by_month <- this_year %>%
      filter(province_state == state) %>%
      group_by(admin2) %>%
      mutate(new_cases=case - lag(case)) %>%
      drop_na(new_cases) %>%
      group_by(month=month(date) + add_months) %>%
      summarize(new_monthly_cases=sum(new_cases))

    yearly_data = rbind(yearly_data, this_state_by_month %>% mutate(state=state))
  }

  return(yearly_data)
}

# function to graph new infections data for a specific year
plot_new_cases_by_month <- function(graph_data, title, events) {
  print(graph_data %>%
    ggplot(aes(x = month, y = new_monthly_cases / 1000, color = factor(state))) +
    geom_line() +
    geom_point() +
    labs(title = title, x = "Month", y = "Cases per month (x 1000)", color = "US State") +
    geom_vline(data = events, aes(xintercept = month), color = "red", linetype = "dashed") +
    geom_text(data = events, aes(x = month, y = Inf, label = label), 
              vjust = 1.5, hjust = 0.5, color = "red", angle = 90) +
    theme_minimal()
  )
}
```


#### 1. What are the trends in daily new cases and deaths globally and in specific regions?

Here we'll compare monthly new case totals for select states

```{r}


# A random list of US states
states <- c("Washington",
            "California",
            "Illinois",
            "Arizona",
            "Ohio",
            "New York",
            "Texas",
            "Florida"
            )


# get data for each year
us_new_cases_2020 <- get_new_cases_by_month('2020')
us_new_cases_2021 <- get_new_cases_by_month('2021')
us_new_cases_2022 <- get_new_cases_by_month('2022')

# make yearly graphs
plot_new_cases_by_month(us_new_cases_2020, "New Cases: 2020")
plot_new_cases_by_month(us_new_cases_2021, "New Cases: 2021")
plot_new_cases_by_month(us_new_cases_2022, "New Cases: 2022 (Jan - April)")


# key events that could cause an impact
events2020 <- data.frame(
  month = c(10),
  label = c("emergency Pfizer vaccine")
)

events2021 <- data.frame(
  month = c(15, 16, 17, 18,  19, 20),
  label = c("vaccine administered", 
            "Delta variant", 
            "anti vaccination misinformation", 
            "Pfizer FDA approval",
            "mandatory federal employees",
            "Younger age vaccines")
)

events2022 <- data.frame(
  month = c(25, 26),
  label = c("Expanded eligibility", 
            "Vaccines for children")

us_new_cases_all_years <- rbind(us_new_cases_2020, us_new_cases_2021, us_new_cases_2022)
plot_new_cases_by_month(us_new_cases_2020, "New Cases: 2020",events2020)
plot_new_cases_by_month(us_new_cases_2021, "New Cases: 2021",events2021)
plot_new_cases_by_month(us_new_cases_2022, "New Cases: 2022",events2022)
plot_new_cases_by_month(us_new_cases_all_years, "New cases by month: 01/2020 - 04/2022")
```

We can see some obvious trends here, especially looking at the whole timeline. The Holiday
Season in 202 saw a pretty big spike, and the same period in 2021 having easily double as
high of a spike. Summer also had a general bump everywhere, and 2021 again being much higher
than 2020.

California had some incredible spikes in the Holiday season, hitting a million new cases
in Dec 2020 and 3 million in Dec 2021. New York, Florida, and Texas
also had big spikes, especially in Summers. Washington remained at one of the
lowest rates throughout the timeline.

Everything trends pretty low at the end of our data in April 2022. Looking at some other
sources, there was a mild peak the following summer, but levels have remained fairly low
since then. We haven't seen any really big surges since the end of 2021.

#### 2. How do different countries compare in terms of COVID-19 cases and deaths?


#### 3. What is the relationship between public health measures (e.g., lockdowns) and the trends in COVID-19 cases?

Here we add some marks for when lockdowns occured in the USA.
```{r}
ld_us <- lockdown %>% 
  filter(country_region == "US") %>% 
  filter(province %in% states) %>% 
  group_by(month=as.numeric(month(date)))

us_new_cases_all_years <- rbind(us_new_cases_2020, us_new_cases_2021, us_new_cases_2022)

us_new_cases_2020 %>% 
    ggplot(aes(x=month,y=new_monthly_cases/1000, color=factor(state))) +
    geom_line() +
    geom_point() +
    labs(title="New cases by month: 2020 with lockdowns", x="month", y="Cases per month (x 1000)", color="US State") +
    geom_vline(data=ld_us, aes(xintercept=month))
    #geom_text(data=ld_us, aes(x=month, y=100 * as.numeric(province), label=province), inherit.aes=FALSE)
```
For the USA, this data is not that interesting because every lockdown happened in March or April of 2020 for states
that had a lockdown. New York seemed to get the most benefit out of these 8. In other states, growth mainly slowed
but still increased slowly after lockdown.

```{r}

```
#### 4. How has the pandemic affected different age groups and genders?

#### 5. What is the impact of vaccination on the spread of COVID-19?
   - plot vaccination rate vs. infection rate
