---
title: "Trends for new Covid-19 infections and impacts of the vaccine"
author: "AB3: Yuxuan Kristin Chen, Christopher Hunter, Sam Malik, Aaron Qiu, Benjamin So"
date: "2024-08-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r include=FALSE}
library(tidyverse)
library(stringr)
library(janitor)
library(lubridate)
library(scales)

# A random list of US states
states <- c("Washington",
            "California",
            "Illinois",
            "Arizona",
            "Ohio",
            "New York",
            "Texas",
            "Florida"
            )

us_i <- clean_names(read_delim("data/us_confirmed.csv"))
us_d <- clean_names(read_delim("data/us_deaths.csv"))
countries <- clean_names(read_delim("data/key-countries-pivoted.csv"))

us_v <- clean_names(read_delim("data/us_state_vaccinations.csv"))

lockdown <-clean_names(read_delim("data/countryLockdowndates.csv"))

# function to get new infection totals per month of a given year
get_new_cases_by_month <- function(data, year) {
  # month scale is absolute from Jan 2020. This gives each month a unique id so
  # we can aggregate year over year data
  add_months <- 12 * (as.numeric(year) - 2020)

  # grab this year's data for the selected states
  this_year <- data %>%
    filter(province_state %in% states) %>%
    filter(str_detect(date, year))

  # we'll add each month to this empty data frame in the loop below
  yearly_data <- data.frame()

  # calculate aggregate new cases per month for each state, and add to the
  # yearly data
  for(state in states) {
    this_state_by_month <- this_year %>%
      filter(province_state == state) %>%
      group_by(admin2) %>%
      mutate(new_cases=case - lag(case)) %>%
      drop_na(new_cases) %>%
      group_by(month=month(date) + add_months) %>%
      summarize(new_monthly_cases=sum(new_cases))

    yearly_data = rbind(yearly_data, this_state_by_month %>% mutate(state=state))
  }

  return(yearly_data)
}

plot_new_cases_by_month <- function(graph_data, title) {
  print(graph_data %>%
    ggplot(aes(x = month, y = new_monthly_cases / 1000, color = factor(state))) +
    geom_line() +
    geom_point() +
    labs(title = title, x = "Month", y = "Cases per month (x 1000)", color = "US State") +
    theme_minimal()
  )
}

plot_new_cases_by_month_event <- function(graph_data, title, events) {
  p <- graph_data %>%
    ggplot(aes(x = month, y = new_monthly_cases / 1000, color = factor(state))) +
    geom_line() +
    geom_point() +
    labs(title = title, x = "Month", y = "Cases per month (x 1000)", color = "US State") +
    geom_vline(data = events,
               aes(xintercept = month), color = "red", linetype = "dashed") +
    geom_text(data = events,
              aes(x = month, y = Inf, label = label),
              vjust = 1.5, hjust = 1.5, color = "red", angle = 90) +
    theme_minimal()

    return(p)
}

```
# Overview {.tabset}

## Abstract

Short summary of our findings

## Introduction

Talk about Covid here

## Data

Our Data was gathered from the following sets available on Kaggle. These are free for public use.

    - https://www.kaggle.com/datasets/gauravduttakiit/covid-19/data
      This is daily updated data on COVID-19 Infections provided by Johns Hopkins University Center for Systems Science and Engineering.

      from this we used 3 data sets:
        - us infections:
          - 5 columns, used: region (county), state, year, number of infections
          - 2,727,072 rows
        - us deaths
          - 5 columns, used: region (county), state, year, number of deaths
          - 2,727,072 rows
        - countries.   : 9 columns, 816 rows

    - https://www.kaggle.com/datasets/paultimothymooney/usa-covid19-vaccinations
      Data about US COVID-19 vaccinations through March 2023

        -  vaccinations:
          - 16 columns, used: month, new_vaccinations
          - 54,628 rows


    - https://www.kaggle.com/datasets/jcyzag/covid19-lockdown-dates-by-country
      Timing of Lockdowns by country

        - lockdown data:
          - 5 columns, used: province, date
          - 307 rows

## Methods

We chose to look at new daily infections/deaths per region for most of our analysis.
This presented a problem because all of the relevant data was only available as
values for running totals over time. We had to do a number of manipulations to the
dataset to create what we wanted to analyze:

- Filter to a limited selection of US states (to avoid cluttering the graphs, and to reduce the amount of data we needed to process)
- Filter cases by year, to be able to compare different periods (seasonal trends, etc)
- Calculate monthly new cases per region by summing daily totals for each month
- Convert time scale to months from start of pandemic. This made it easier to take data for each year and combine it to see long term trends.
- Aggregate monthly totals for each region into the appropriate state

We used a few functions to automate this process, and also to do some of the
plotting, in cases where we were analyzing different sections of similar data.

We eliminated some null fields, such as new cases on the first of the month

## Findings

Conclusions here

## Summary

Summary here

# Research Questions {.tabset}

## US Covid Trends

### 1. What are the trends in monthly new cases and deaths for US states?

We'll look at a few semi-random US states here and compare rates

#### Comparing infections

```{r}

# get data for each year
us_new_cases_2020 <- get_new_cases_by_month(us_i, '2020')
us_new_cases_2021 <- get_new_cases_by_month(us_i, '2021')
us_new_cases_2022 <- get_new_cases_by_month(us_i, '2022')

# make yearly graphs
plot_new_cases_by_month(us_new_cases_2020, "New Cases: 2020")
plot_new_cases_by_month(us_new_cases_2021, "New Cases: 2021")
plot_new_cases_by_month(us_new_cases_2022, "New Cases: 2022 (Jan - April)")

```

We can see some obvious trends here, especially looking at the whole timeline. The Holiday
Season in 202 saw a pretty big spike, and the same period in 2021 having easily double as
high of a spike. Summer also had a general bump everywhere, and 2021 again being much higher
than 2020.

California had some incredible spikes in the Holiday season, hitting a million new cases
in Dec 2020 and 3 million in Dec 2021. New York, Florida, and Texas
also had big spikes, especially in Summers. Washington remained at one of the
lowest rates throughout the timeline.

Everything trends pretty low at the end of our data in April 2022. Looking at some other
sources, there was a mild peak the following summer, but levels have remained fairly low
since then. We haven't seen any really big surges since the end of 2021.

Conclusion: Infections tended to have a peak in summer and a higher peak during the holiday
season. Later years had more overall infections than earlier years. Very large peaks were
not seen after early 2022

#### Comparing deaths

```{r}
# get data for each year
us_new_deaths_2020 <- get_new_cases_by_month(us_d, '2020')
us_new_deaths_2021 <- get_new_cases_by_month(us_d, '2021')
us_new_deaths_2022 <- get_new_cases_by_month(us_d, '2022')

# make yearly graphs
plot_new_cases_by_month(us_new_deaths_2020, "New Deaths: 2020")
plot_new_cases_by_month(us_new_deaths_2021, "New Deaths: 2021")
plot_new_cases_by_month(us_new_deaths_2022, "New Deaths: 2022 (Jan - April)")
```

We see similar trend lines for each year, though there is a very large initial
spike for New York. Also deaths tend to fall away at a slower rate than infections
for spikes in the same time period. Again, large population centers seem to be
hit hardest.

Conclusion: Death rates follow similar patterns to infection rates, but were much
higher in the very early period of the pandemic. Dense population centers were more
impacted by deaths.

#### Infections vs. Deaths
```{r}
us_new_cases_all_years <- rbind(us_new_cases_2020, us_new_cases_2021, us_new_cases_2022)
plot_new_cases_by_month(us_new_cases_all_years, "New cases by month: 01/2020 - 04/2022")

us_new_deaths_all_years <- rbind(us_new_deaths_2020, us_new_deaths_2021, us_new_deaths_2022)
plot_new_cases_by_month(us_new_deaths_all_years, "New deaths by month: 01/2020 - 04/2022")
```
We can see clearly that deaths were much more severe in the earlier months of the
pandemic. Very large spikes early on, especially in New York, but we see that even
the biggest infection spikes in 2021 had relatively fewer deaths associated with
the same time periods.

Conclusion: One impact of the vaccine was that the relative amount of deaths from
covid decreased considerably once the vaccine was in widespread use, even as
infections increased

## Key Countries

### 2. How do different countries compare in terms of COVID-19 cases?
```{r}
countries_pivoted <- countries %>%
  pivot_longer(!date, names_to="country", values_to="cases")

countries_pivoted %>%
  group_by(date) %>%
  ggplot(aes(x=date, y=cases/1000000, color=factor(country))) +
  geom_path() +
  labs(title="Infections by Country", x="Date", y="Cases (millions)", color="Country")
```

We can see an obvious trend that infections in the US are far higher than the other
key countries throughout the timeline, though the US has proportionally more population
than many of the countries here. China is an outlier in that their reported numbers
are very low for their large population.

Additionally we can see that total cases actually increased at a much higher rate
starting in early 2022. People tend to regard the first two years as the most
impactful time, but we can see that infections actually have skyrocketed after that.

Conclusion: We can guess that in 2022 and later, many citizens felt protected by the
vaccine and were less concerned about contracting covid, so they may have begun to
engage in more social behaviors which allowed the virus to spread more rapidly.

## Lockdowns

### 3. What is the relationship between public health measures (e.g., lockdowns) and the trends in COVID-19 cases?

Here we add some marks for when lockdowns occured in the USA.
```{r fig.width=10}
ld_us <- lockdown %>%
  filter(country_region == "US") %>%
  filter(province %in% states) %>%
  mutate(date=dmy(date)) %>%
  group_by(month=as.numeric(yday(date) / 30) + 1 ) %>%
  reframe(month, label=province)

early_2020_cases <-us_new_cases_2020 %>%
  filter(month < 6, month > 2)

plot_new_cases_by_month_event(early_2020_cases, "New Cases: 2020: March - May", ld_us)
```

For the USA, this data is not that interesting because every lockdown happened in
March or April of 2020 for states that had a lockdown. New York seemed to get the
most benefit out of these 8. In other states, growth mainly slowed
but still increased slowly after lockdown.

Data for other countries is pretty similar. Lockdowns all began roughly in the
first 4 months of 2020, so it's difficult to draw any meaningful conclusions about
their effects, especially since infection rates were relatively low for many areas
in the very early part of the pandemic.

Additionally, many lockdowns were lifted, put back in place, and lifted again
several times over, and the timing of this varied widely between countries and
in different US states, so there is often not a direct timeline equivalent between
different areas that would be useful to compare.

Conclusion: Lockdowns tended to have a positive impact in densely populated areas,
but aside from that we can’t draw any major conclusions from this data, partly
because they all happened at a similar time. Also there is no way to compare to
what infection rates would have been without a lockdown.

## Impact of Vaccines

### 4. What is the impact of vaccination on the spread of COVID-19?
We can map a few key events in the vaccination timeline to the rate of infections

```{r fig.width=10}
# key events that could cause an impact
events2020 <- data.frame(
  month = c(10),
  label = c("emergency Pfizer vaccine")
)

events2021 <- data.frame(
  month = c(12, 16, 17, 18,  19, 20),
  label = c("vaccine administered",
            "Delta variant",
            "anti vaccination misinformation",
            "Pfizer FDA approval",
            "mandatory federal employees",
            "Younger age vaccines")
)

events2022 <- data.frame(
  month = c(25, 26),
  label = c("Expanded eligibility",
            "Vaccines for children")
)

combined_events <- bind_rows(events2020, events2021, events2022)
plot_new_cases_by_month_event(us_new_cases_all_years, "New cases by month: 01/2020 - 04/2022 with vaccine events", combined_events)
```
The introduction of emergency-use authorization for the Pfizer vaccine and
subsequent vaccine administration led to a significant initial reduction in case
numbers. However, later events such as the Delta variant and misinformation
surrounding vaccines contributed to subsequent spikes in cases.

The expanded eligibility and the approval of vaccines for younger age groups
appear to coincide with the gradual decline in cases after the peak of the Delta
variant. This suggests that broadening vaccine coverage was crucial in controlling
the spread of the virus.

Across the states shown, the trends are remarkably similar, with all experiencing
peaks at similar times, reflecting the nationwide impact of variants and\
vaccine-related policies.

Conclusion: We can see correlations between infection rates and when various
vaccine developments happened.

## Vax rates vs. infections

### 5. How does the rate of vaccinations compare to the infection rate?

```{r fig.width=10, echo=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
us_v_us <- us_v %>%
  filter(location == "United States") %>%
  arrange(date) %>%
  fill(total_vaccinations, .direction = "down") %>%
  mutate(new_vaccinations = total_vaccinations - lag(total_vaccinations, default = 0)) %>%
  mutate(month = floor_date(date, "month"))  %>%
  group_by(month) %>%
  summarise(new_vaccinations = sum(new_vaccinations, na.rm = TRUE))

us_i_us <- us_i %>%
  group_by(date) %>%
  summarise(total_cases = sum(case, na.rm = TRUE)) %>%
  mutate(new_cases = total_cases - lag(total_cases, default = 0)) %>%
  mutate(month = floor_date(date, "month")) %>%
  group_by(month) %>%
  summarise(new_cases = sum(new_cases, na.rm = TRUE))

combined_data <- us_v_us %>%
  full_join(us_i_us, by = "month")

ggplot(combined_data, aes(x = month)) +
    geom_line(aes(y = new_vaccinations, color = "New Vaccinations"), linewidth = 1) +
    geom_line(aes(y = new_cases, color = "New Cases"), linewidth = 1) +
    labs(title = "Monthly New Vaccinations and New Cases in the US",
         x = "Month",
         y = "Count",
         color = "Legend") +
    scale_x_date(date_labels = "%Y-%m", date_breaks = "1 month") +
    scale_y_continuous(labels = label_number()) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          plot.title = element_text(hjust = 0.5, face = "bold"))
```

