---
title: "final_data"
author: "Chris Hunter"
date: "2024-08-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(stringr)
library(janitor)
library(lubridate)
us_i <- read_delim("infections/data/us_confirmed.csv")
us_d <- read_delim("infections/data/us_deaths.csv")
us_i <- clean_names(us_i)
us_d <- clean_names(us_d)

#vax <- read_delim("vax/us_state_vaccinations.csv")
#ld <- read_delim("lockdown/countryLockdowndates.csv")
```



#### 1. What are the trends in daily new cases and deaths globally and in specific regions?
compare monthly case totals in 2022 for select states
```{r}
states <- c("Washington", "California", "Colorado", "Arizona", "Ohio", "Alabama", "Missouri", "South Dakota", "New York", "Texas", "Florida", "Kentucky" )

us_i_2022 <- us_i %>%
  filter(province_state %in% states) %>% 
  filter(str_detect(date, '2022'))

us_d_2022 <- us_d %>%
  filter(province_state %in% states) %>% 
  filter(str_detect(date, '2022'))


us_i_2022 %>% 
  group_by(month = month(date)) %>% 
  summarize(monthly_cases = sum(case))
  
```

#### 2. How do different countries compare in terms of COVID-19 cases and deaths?
      - plot 
    
#### 3. What is the relationship between public health measures (e.g., lockdowns) and the trends in COVID-19 cases?
  - plot infection rate vs. lockdown times

#### 4. How has the pandemic affected different age groups and genders?

#### 5. What is the impact of vaccination on the spread of COVID-19?
   - plot vaccination rate vs. infection rate
